
---
title: "Data Wrangling with dplyr and tidyr"
output: html_document
---

## Introduction

This document demonstrates various data wrangling operations using the `dplyr` and `tidyr` packages in R. The operations include filtering, selecting, adding new columns, summarizing, sorting, handling missing values, merging datasets, and reshaping data.

## Step 1: Create Synthetic Data

We begin by creating a synthetic dataset using the `tibble` function.

```{r}
library(dplyr)
library(tidyr)

# Create synthetic data
set.seed(123)
df <- tibble(
  ID = 1:10,
  Name = c("Alice", "Bob", "Charlie", "David", "Eva", "Frank", "Grace", "Hannah", "Ian", "Jane"),
  Age = sample(18:60, 10, replace = TRUE),
  Gender = sample(c("Male", "Female"), 10, replace = TRUE),
  Salary = sample(30000:80000, 10, replace = TRUE)
)

# Print the original dataframe
df
```

## Step 2: Data Wrangling Operations

### 2.1 Filter rows where Age > 30

We filter the rows where `Age` is greater than 30.

```{r}
# Filter rows where Age > 30
filtered_df <- df %>% filter(Age > 30)
filtered_df
```

### 2.2 Select specific columns (Name and Age)

We select only the `Name` and `Age` columns from the dataset.

```{r}
# Select columns 'Name' and 'Age'
selected_columns <- df %>% select(Name, Age)
selected_columns
```

### 2.3 Add a new column 'AgeGroup'

We create a new column `AgeGroup` that categorizes individuals as "Adult" or "Child" based on their age.

```{r}
# Add a new column 'AgeGroup'
df <- df %>% mutate(AgeGroup = if_else(Age > 18, "Adult", "Child"))
df
```

### 2.4 Summarize by grouping Gender and calculating mean Age

We group the data by `Gender` and calculate the mean age for each group.

```{r}
# Summarize by Gender and calculate mean Age
summary <- df %>% group_by(Gender) %>% summarize(mean_age = mean(Age, na.rm = TRUE))
summary
```

### 2.5 Sort by Age in descending order

We sort the dataset by `Age` in descending order.

```{r}
# Sort by Age in descending order
sorted_df <- df %>% arrange(desc(Age))
sorted_df
```

### 2.6 Handle missing values in the Salary column

We introduce a missing value (`NA`) in the `Salary` column and then fill it with the mean salary of the non-missing values.

```{r}
# Handle missing values in the Salary column
df$Salary[2] <- NA  # Introduce a missing value
df <- df %>% mutate(Salary = if_else(is.na(Salary), mean(Salary, na.rm = TRUE), Salary))
df
```

### 2.7 Merge with another DataFrame

We merge the original dataframe `df` with another dataframe `df2` containing `Department` information. We use a left join on the `ID` column.

```{r}
# Merge with another DataFrame
df2 <- tibble(
  ID = c(1, 2, 3, 4, 5),
  Department = c("HR", "Finance", "IT", "Marketing", "Sales")
)
merged_df <- df %>% left_join(df2, by = "ID")
merged_df
```

### 2.8 Pivot DataFrame

We pivot the dataset to widen it, spreading `Salary` values across new columns based on the `AgeGroup`.

```{r}
# Pivot the DataFrame
pivot_df <- df %>% pivot_wider(names_from = AgeGroup, values_from = Salary, values_fn = mean)
pivot_df
```

## Conclusion

This document demonstrated several essential data wrangling operations in R using the `dplyr` and `tidyr` packages, including filtering, selecting, transforming, summarizing, and reshaping data.
